Potential competition in alpine bumble bees is mitigated by imperfect trait-matching and subpartitioning within floral morphotypes

### 2021-03-09 ms_2/ms_2_v2.Rmd
did some writing and added some references

Bumble bees (Hymenoptera: *Bombus* spp.) have become classic model organisms for the study of competition and species coexistence due to their broad and overlapping distributions and their gross similarity in morphology, behavior, and dietary niche [@Plowright1984-lc; @Goulson2008-zy]. In this role, however, they have proven somewhat enigmatic. On the one hand, there is evidence that resource limitation can be a significant constraint on local bumble communities [@Heinrich1976-wc; @Bowers1985-df], and certain bumble bee communities have been shown to be over-dispersed in terms of tongue length (a matching trait corresponding to floral corolla depth) or phylogeny (understood to encompass unmeasured trait variation), suggesting that competition functions as a community assembly process [@Inouye1977-kl; @Pellissier2013-pk]. Nevertheless, bumble bee communities with an apparent excess of species richness and deficit of trait (or phylogenetic) dispersion are commonly observed [@Ranta1981-vx; @Goulson2008-zy]. Teasing apart the still-obscure mechanisms of bumble bee coexistence and niche partitioning remains a salient problem from a basic ecological perspective, but it has also become an urgent question in the context of applied conservation. Bumble bees are cold-adapted species [@Heinrich1994-cv] associated mainly with mountain habitats, where they serve as the principal pollinators of many alpine plants [@Williams1998-xi; @Goulson2010-lo]. Warming climates have already caused measurable upslope shifts in bumble bees and their floral hosts [@Ploquin2013-da; @Kerr2015-dg; @Pyke2016-ll; @Fourcade2019-ct; @Soroye2020-zb; @Marshall2020-pc], and these are predicted to cause declines or even extinction in many bumble bees species [@Soroye2020-zb]. One mechanism by which climate change is expected to affect bumble bees is the introduction of novel interspecific competition driven by the upslope advance of lowland bumble bee species into the elevational distributions of their highland counterparts. Thus, the speculative question of how competition has acted in the historic assembly of bumble bee communities has become an unfolding process observable over ecological time scales and instrumental in the future viability of bumble bee species and their floral mutualists.

@@

Kugler morphotype classifications of the plants in our study system were accessed via the BIOFLOR database (Klotz, Kühn, and Durka 2002). Morphotypes were simplified to the primary classes numbered 0-10: 0 = not applicable (includes wind-pollinated flora such as grasses), 1 = disc- and bowl-shaped flowers, 2 = funnel flowers, 3 = bell-shaped flowers, 4 = stalk-disc flowers, 5 = lip flowers, 6 = flag flowers, 7 = flower heads, 8 = spike flowers, 9 = brush flowers, 10 = trap flowers **(Figure 1)**. In a few cases, a plant species in our data set was not found in the BIOFLOR database. In most cases, morphotypes are constant at the genus level, and missing morphotypes were added based on the morphotypes found for congeners. The genus *Gentiana* consisted of both group 2 and group 4 representatives, so the missing species *Gentiana aspera* and *Gentiana ciliata* were classified based on visual comparison with their classified congeners. In the case of the genus *Phyteuma*, we followed Neumayer and Paulus (1999) and classified it as a group 7 flower head. Morphotypes 9 and 10 were extremely rare in our study system and were therefore omitted from all analyses. Plants belonging to morphotype 0 were typically not recorded during floral surveys due to their lack of conspicuous floral displays, and they were rarely visited by bumble bees, so this morphotype was also omitted from all analyses.

@@

Bumble bee tongue-length data were obtained from the literature values compiled and summarized by Arbetman et al. (2017). Values for *B. gerstaeckeri* and *B. mendax*, which were not included in Arbetman et al. (2017), were obtained from Obeso (1992) and Durieux (2000), respectively. We then binned bumble bee species into three tongue-length classes: short (< 7.5 mm), medium (7.5—10.0 mm), and long (>= 10.0 mm). Short-tongued species included *B. jonellus*, *B. lapidarius*, *B. pratorum*, *B. psithyrus*-group, *B. pyrenaeus*, and *B. terrestris/lucorum*. Medium-tongued species included *B. humilis*, *B. hypnorum*, *B. monticola*, *B. mucidus*, *B. pascuorum*, and *B. wurflenii*. Long-tongued species included *B. gerstaeckeri*, *B. hortorum*, and *B. mendax*.

@@

All analyses were conducted in `R` `r citep(citation())`. Data handling and visualization was performed with the `tidyverse` suite of R packages `r citep(citation("tidyverse"))`. Bayesian multilevel models were implemented in Stan (Stan Development Team 2021) via the package `brms` `r citep(citation("brms"))` using default priors, and visualizations were prepared with the packages `tidybayes` `r citep(citation("tidybayes"))` and `ggdist` `r citep(citation("ggdist"))`. Ordination analysis and permutation testing were performed with the package `vegan` `r citep(citation("vegan"))` and visualized using the packages `ggplot2` `vegan` `r citep(citation("ggplot2"))` and `ggvegan` (Simpson 2019). Hierarchical generalized additive models (HGAMs) were implemented with the package `mgcv` [@Wood_2017] and visualized with the packages `mgcViz` [@Fasiolo2018aa] and `tidymv` [@Coretta-aa2021]. All R code used in this analysis, together with full model output and diagnostics, is available in Appendix X.

@@

To investigate elevational patterns of *within-site* niche overlap, we modeled niche overlap as a function of level (morphotype, family, genus) and elevation using HGAMs. First, we calculated community mean niche overlap by taking the mean niche overlap (within level) across all species pairs for each site. Level was included as both a fixed intercept effect and a smoothing factor in conjunction with elevation, following the type "I" model form described by @Pedersen2019-kf. To account for the proportional nature of our data, we used a beta distribution family with a logit link function.

@@

Disc (group 1) flowers were visited at similar rates across tongue length classes **(Figure 2)**. At the species level, disc flowers were visited at relatively high rates by the short-tongued B. terrestris/lucorum and the medium-tongued Bombus monticola **(Figure 3)**. Visitation to funnel (group 2) flowers was dominated by long-tongued species, an effect driven entirely by the Aconitum (group 2) specialist B. gerstaeckeri. Bell (group 3) flowers were visited at a higher rate by short-tongued species than by long-tongued species, with medium-tongued species exhibiting intermediate visitation rates. At the species-level, bell flowers were visited at highest probability by the short-tongued species B. jonellus, B. pyrenaeus, and B. soroeensis, and by the medium-tongued species B. hypnorum and B. monticola. Stalk-disc (group 4) flowers were visited at similar rates across all tongue length classes, and there was also broad overlap in visitation rates at the species level. Lip (group 5) flowers were also visited at similar rates across tongue length classes, but at the species level B. pascuorum exhibited an uniquely high bias toward lip flowers. Flag (group 6) flowers were visited at a higher rate by medium-tongued species than by short-tongued species, with long-tongued species exhibiting intermediate visitation rates. Flag flowers generated strong separation in visitation rate at the species level, receiving very high visitation from the medium-tongued B. pascuorum and B. mucidus, moderately high visitation from the short-tongued B. terrestris/lucorum and B. lapidarius, the medium-tongued B. monticola, and the long-tongued B. mendax and B. hortorum, and low visitation by the short-tongued B. soroeensis, B. pyrenaeus, B. psithyrus, B. pratorum, and B. jonellus, the medium-tongued B. hypnorum, and the long-tongued B. gerstaeckeri. Models 2 and 3 both explained approximately 31% of total variance (97.5% CI = 30–32%).

@@

The mystery of bumble bee species coexistence hinges on the relationship between species-level floral preferences, trait-matching, and the emergent phenomenon of community-level patterns of diet overlap. In this study, we show that while tongue length is a relatively strong predictor of species-level floral preferences, it is at best a weak predictor of diet overlap. This is due in part to divergent preference of bumble bees with similar tongue lengths toward certain floral morphotypes. Perhaps more importantly, though, diet overlap at the level of floral genus tended to be about half as strong as diet overlap at the level of floral morphotype, indicating that bumble bees with similar morphological affinities, likely associated with tongue length, subdivide floral morphotypes at the genus- and species-level, reducing the potential for competition.

@@

and resulting patterns of dietary overlap. In this study, we demonstrate that while bumble bees are, indeed, remarkably generalized in their foraging, they are highly discriminating generalists, over-representing certain floral taxa and morphotypes and ignoring others. These patterns of floral preference are related to bumble bee tongue length, but there exists considerable variation in floral preference among species with similar tongue lengths, which may serve to relax competition between them.

In the cases of disc and stalk-disc flowers visitation rates were uniformly low across all tongue length classes. The opposite was true of lip flowers, which were visited at similarly high rates by all tongue length classes. Bell flowers and head flowers were visited chiefly by short-tongued bees, consistent with the shallow corollas of these morphotypes. Conversely, funnel flowers, with their deep corollas were visited chiefly by long-tongued bees, though this effect was driven disproportionately by *B. gerstaeckeri*, which specializes on the funnel flowers of the genus *Aconitum* [@Ponchau2006-af]. Flag flowers were visited at the highest rate by medium-tongued species, and especially by *B. pascuorum* and *B. mucidus*.

If not tongue length, what else could explain the variation in floral morphotype preference witin tongue length classes?

When bumble bee visitation is pooled across all species, it is clear that lip, flag, and head flowers dominate the floral community in terms of conditional visitation probability. Patterns of pooled visitation are, of course, driven disproportionately by the behavior of the most abundant bumble bee species, which in our system were *B. pascuorum*, *B. pratorum*, *B. soroeensis*, *B. terrestris/lucorum*, and *B. wurflenii* (Sponsler et al., submitted). Visitation to lip and flag flowers was driven especially by the medium-tongued *B. pascuorum*, the most abundant bumble bee in our study system, and secondarily by the medium-tongued *B. wurflenii*. Visitation to head flowers was driven mainly by the short-tongued *B. soroeensis*.

The influence of tongue length on visitation bias was pronounced for some morphotypes but not for others. In the cases of disc and stalk-disc flowers visitation rates were uniformly low across all tongue length classes. The opposite was true of lip flowers, which were visited at similarly high rates by all tongue length classes. Bell flowers and head flowers were visited chiefly by short-tongued bees, consistent with the shallow corollas of these morphotypes. Conversely, funnel flowers, with their deep corollas were visited chiefly by long-tongued bees, though this effect was driven disproportionately by *B. gerstaeckeri*, which specializes on the funnel flowers of the genus *Aconitum* [@Ponchau2006-af]. Flag flowers were visited at the highest rate by medium-tongued species, and especially by *B. pascuorum* and *B. mucidus*.

While the variation in floral bias across tongue length classes we observed largely comports with the established relationship between tongue length and corolla depth, variation in bias within tongue length classes demonstrates the insufficiency of simple morphological matching to explain bumble bee foraging patterns. Nectar robbing is one mechanism by which the relationship between tongue length and corolla depth might be relaxed, but its influence in our data is unclear. Of the bumble bees in our study system, *B. terrestris/lucorum* and especially *B. wurflenii* are known to engage frequently in nectar robbing [@Brian1954-cy]. In the case of *B. wurflenii*, however, its morphotype affinities did not differ markedly from those of other medium-tongued species, and it may be that the condition of medium-tongue length alone confers enough to foraging flexibility in our study system to dampen the divergent effects of nectar robbing. Nectar robbing might, however, account for the deviation of *B. terrestris/lucorum* from most other short-tongued species with respect to flag flowers. With the exception of *B. lapidarius*, all short-tongued species besides *B. terrestris/lucorum* eschewed flag flowers, while *B. terrestris/lucorum* visited them at rates comparable to medium- and long-tongued species. Indeed, B. terrestris/lucorum was, on the whole, an idiosyncratic forager, also demonstrating a relatively high affinity for the disc flowers that were largely ignored by other species. Perhaps the most striking divergence of floral bias within tongue length class, though, was the divisive effect of bell flowers on short-tongued bumble bees. *B. pyrenaeus*, *B. soroeensis*, *B. pratorum*, and *B. jonellus* visited bell flowers at high rates, while *B. terrestris/lucorum*, *B. psithyrus*, and *B. lapidarius* visited them relatively rarely. A similar split among the short-tongued species occurred with respect to head flowers, which were visited at high rates by *B. soroeensis* and *B. psithyrus* but at relatively low rates by other short-tongued species. There are no obvious morphological distinctions within short-tongued bumble bees that would account for these divergent affinities, and ecological drivers such as differences in spatial or temporal activity patterns are more likely explanations.

Our finding that bumble bee diet ordination produces a primary axis corresponding to tongue length (and associated with affinity for legumes) and a secondary axis describing the dispersion of short-tongued species is consistent with studies in Poland [@Goulson2008-oe], France [@Durieux2000-uk], and England [@Goulson2005-nz] that analyzed bumble bee diet by similar methods, suggesting that this may be a general pattern among bumble bee communities, at least within temperate Europe. While our test of beta-dispersion did not reveal significant differences between tongue length classes, the spread of short-tongued bees along the secondary axis of our ordination, together with their divergence with respect to visitation rates on bell and head flowers described above, suggests that resource partitioning may be more pronounced among short-tongued bumble bees than among their medium- and long-tongued counterparts.

It is tempting to invoke interspecific competition as a cause of the apparent niche partitioning among short-tongued bumble bees, in agreement with previous studies that have demonstrated the potential for severe competition between bumble bees of similar tongue length (e.g. Heinrich 1976; Inouye 1978; Pyke 1982). But the importance of competition in bumble bee communities is not universal (e.g. Ranta and Vepsäläinen 1981; Williams 1989), and in the rich wildflower meadows of our (sub)alpine system, bumble bee populations are more likely limited by harsh abiotic conditions than by food availablility. Alternatively, it may be that short-tongued bees in our study system differed from one another in their affinities for specific habitats (Schneller et al. 2014) or spatial configurations of resource patches (Sowig 1989), and their divergent morphological biases arose secondarily from these effects.

## Diet overlap

We analyze community mean niche overlap as an indication of diffuse competition [@Pianka1974-qk; MacArthur1972-qz]

To account for this inconsistency, @Ranta1981-vx proposed that bumble bees communities are governed by a spatiotemporally heterogeneous mosaic of competitive conditions, such that competitive inequalities between species are sufficiently variable to allow more species to coexist than predicted by trait-based competition models. A complementary explanation, based on the core/satellite species hypothesis [@Hanski1982-hs], is that competitive forces are significant only among regionally abundant and widespread “core species”, and the effects of competition are more pronounced in the structure of regional species pools than in that of local assemblages [@Hanski1982-ie]. In some cases, competition may simply be precluded by a saturating abundance of floral resources [@Williams1989-vr]. Range shifts in bumble bees have already been documented [@Ploquin2013-da; @Kerr2015-dg; @Pyke2016-ll; @Fourcade2019-ct; @Soroye2020-zb; @Marshall2020-pc], and the introduction or extinction of species from bumble bee communities, with corresponding addition and release of competitive pressure, can alter both the floral visitation patterns of bumble bees and, consequently, the pollination biology of their host plants [@Brosi2013-mu; @Ishii2013-sm].

We see that community-mean diet overlap, indicative of the potential for diffuse competition, generally declines with elevation [@Pianka1974-qk].

"there  is  still  considerable  scope  for  niche  differentiation  among  species  with  tonguesthat  are  near-identical  in  length." @Goulson2008-zy

# CONCLUSIONS

For bumble bees, floral bias is a labile product of innate biases, the soft constraints of morphological compatibility, and the dynamic ecological context in which foraging occurs. By conditioning bumble bee visitation data upon concurrent floral abundance data across spatially and temporally replicated samples, we demonstrate the strength bumble bee floral bias, its functional underpinnings, its variation across species, and its lability within species in response to environmental variation. Such tuning of floral bias has obvious adaptive significance for bumble bees, which are characteristically — and most likely primordially — associated with mountain habitats (Williams 1985; Hines 2008), where steep temperature gradients and short growing seasons generate extreme spatial and temporal floral turnover.

The aim of our study was to build an analysis of bumble bee diet overlap from the ground up, as it were, beginning with a description of floral preferences, modeling these preferences with respect to bumble bee and floral traits, and

Our study begins at the level of bumble bee floral visitation patterns, which, when conditioned on the neutral effects of abundance, can be interpreted as floral preferences. These floral preferences, together with potential effects of spatial and temporal co-occurrence,

>>>>

Bumble bees (Hymenoptera: *Bombus* spp.) have become classic model organisms for the study of competition and species coexistence due to their broad and overlapping distributions and their gross similarity in morphology, behavior, and dietary niche [@Plowright1984-lc; @Goulson2008-zy]. In this role, however, they have proven somewhat enigmatic. On the one hand, there is evidence that resource limitation can be a significant constraint on local bumble communities [@Heinrich1976-wc; @Bowers1985-df], and certain bumble bee communities have been shown to be over-dispersed in terms of tongue length (a matching trait corresponding to floral corolla depth) or phylogeny (understood to encompass unmeasured trait variation), suggesting that competition functions as a community assembly process [@Inouye1977-kl; @Pellissier2013-pk]. Nevertheless, bumble bee communities with an apparent excess of species richness and deficit of trait (or phylogenetic) dispersion are commonly observed [@Ranta1981-vx; @Goulson2008-zy]. To account for this inconsistency, @Ranta1981-vx proposed that bumble bees communities are governed by a spatiotemporally heterogeneous mosaic of competitive conditions, such that competitive inequalities between species are sufficiently variable to allow more species to coexist than predicted by trait-based competition models. A complementary explanation, based on the core/satellite species hypothesis [@Hanski1982-hs], is that competitive forces are significant only among regionally abundant and widespread “core species”, and the effects of competition are more pronounced in the structure of regional species pools than in that of local assemblages [@Hanski1982-ie]. In some cases, competition may simply be precluded by a saturating abundance of floral resources [@Williams1989-vr].

Teasing apart the still-obscure mechanisms of bumble bee coexistence and niche partitioning remains a salient problem from a basic ecological perspective, but it has also become an urgent question in the context of applied conservation. Bumble bees are cold-adapted species [@Heinrich1994-cv] associated mainly with mountain habitats, where they serve as the principal pollinators of many alpine plants [@Williams1998-xi; @Goulson2010-lo]. Warming climates have already caused measurable upslope shifts in bumble bees and their floral hosts [@Ploquin2013-da; @Kerr2015-dg; @Pyke2016-ll; @Fourcade2019-ct; @Soroye2020-zb; @Marshall2020-pc], and these are predicted to cause declines in many bumble bees species [@Soroye2020-zb]. Moreover, the introduction or extinction of species from bumble bee communities, with corresponding addition and release of competitive pressure, can alter both the floral visitation patterns of bumble bees and, consequently, the pollination biology of their host plants [@Brosi2013-mu; @Ishii2013-sm]. Thus, the speculative question of how competition has acted in the historic assembly of bumble bee communities has become an unfolding process observable over ecological time scales and instrumental in the future viability of bumble bee species and their floral mutualists.

@@

Kugler morphotype classifications of the plants in our study system were accessed via the BIOFLOR database [@Klotz2002-uc]. Morphotypes were simplified to the primary classes numbered 0-10: 0 = not applicable (includes wind-pollinated flora such as grasses), 1 = disc- and bowl-shaped flowers, 2 = funnel flowers, 3 = bell-shaped flowers, 4 = stalk-disc flowers, 5 = lip flowers, 6 = flag flowers, 7 = flower heads, 8 = spike flowers, 9 = brush flowers, 10 = trap flowers **(Figure 1)**. In a few cases, a plant species in our data set was not found in the BIOFLOR database. In most cases, morphotypes are constant at the genus level, and missing morphotypes were added based on the morphotypes found for congeners. The genus *Gentiana* consisted of both group 2 and group 4 representatives, so the missing species *Gentiana aspera* and *Gentiana ciliata* were classified based on visual comparison with their classified congeners. In the case of the genus *Phyteuma*, we followed @Neumayer1999-il and classified it as a group 7 flower head. Morphotypes 9 and 10 were extremely rare in our study system and were therefore omitted from all analyses. Plants belonging to morphotype 0 were typically not recorded during floral surveys due to their lack of conspicuous floral displays, and they were rarely visited by bumble bees, so this morphotype was also omitted from all analyses.

@@

Bumble bee tongue-length data were obtained from the literature values compiled and summarized by @Arbetman2017-xj. Values for *B. gerstaeckeri* and *B. mendax*, which were not included in @Arbetman2017-xj, were obtained from @Obeso1992-yp and @Durieux2000-uk, respectively. We then binned bumble bee species into three tongue-length classes: short (< 7.5 mm), medium (7.5—10.0 mm), and long (>= 10.0 mm). Short-tongued species included *B. jonellus*, *B. lapidarius*, *B. pratorum*, *B. psithyrus*-group, *B. pyrenaeus*, and *B. terrestris/lucorum*. Medium-tongued species included *B. humilis*, *B. hypnorum*, *B. monticola*, *B. mucidus*, *B. pascuorum*, and *B. wurflenii*. Long-tongued species included *B. gerstaeckeri*, *B. hortorum*, and *B. mendax*.

@@

All analyses were conducted in `R` `r citep(citation())`. Data handling and visualization was performed with the `tidyverse` suite of R packages [@Wickham2019-aa]. Bayesian multilevel models were implemented in Stan [@Stan2020aa] via the package `brms` [@Burkner2017-aa] using default priors, and visualizations were prepared with the packages `tidybayes` `r citep(citation("tidybayes"))` and `ggdist` `r citep(citation("ggdist"))`. Ordination analysis and permutation testing were performed with the package `vegan` `r citep(citation("vegan"))` and visualized using the packages `ggplot2` [@Wickham2016-aa] and `ggvegan` [@Simpson2019aa]. Hierarchical generalized additive models (HGAMs) were implemented with the package `mgcv` [@Wood_2017] and visualized with the packages `mgcViz` [@Fasiolo2018aa] and `tidymv` [@Coretta-aa2021]. All R code used in this analysis, together with full model output and diagnostics, is available in Appendix X.

@@

To investigate elevational patterns of *within-site* niche overlap, we modeled community-mean niche overlap as a function of level (morphotype, family, genus) and elevation using HGAMs. Community mean niche overlap was calculated by taking the mean niche overlap (within level) across all species pairs for each site. Level was included as both a fixed intercept effect and a smoothing factor in conjunction with elevation, following the type "I" model form described by @Pedersen2019-kf. To account for the proportional nature of our data, we used a beta distribution family with a logit link function.

@@

Disc (group 1) flowers were visited at similar rates across tongue length classes **(Figure 2)**. At the species level, disc flowers were visited at relatively high rates by the short-tongued *B. terrestris/lucorum* and the medium-tongued *Bombus monticola* **(Figure 3)**. Visitation to funnel (group 2) flowers was dominated by long-tongued species, an effect driven entirely by the *Aconitum* (group 2) specialist *B. gerstaeckeri*. Bell (group 3) flowers were visited at a higher rate by short-tongued species than by long-tongued species, with medium-tongued species exhibiting intermediate visitation rates. At the species-level, bell flowers were visited at highest probability by the short-tongued species *B. jonellus*, *B. pyrenaeus*, and *B. soroeensis*, and by the medium-tongued species *B. hypnorum* and *B. monticola*. Stalk-disc (group 4) flowers were visited at similar rates across all tongue length classes, and there was also broad overlap in visitation rates at the species level. Lip (group 5) flowers were also visited at similar rates across tongue length classes, but at the species level *B. pascuorum* exhibited an uniquely high bias toward lip flowers. Flag (group 6) flowers were visited at a higher rate by medium-tongued species than by short-tongued species, with long-tongued species exhibiting intermediate visitation rates. Flag flowers generated strong separation in visitation rate at the species level, receiving very high visitation from the medium-tongued *B. pascuorum* and *B. mucidus*, moderately high visitation from the short-tongued *B. terrestris/lucorum* and *B. lapidarius*, the medium-tongued *B. monticola*, and the long-tongued *B. mendax* and *B. hortorum*, and low visitation by the short-tongued *B. soroeensis*, *B. pyrenaeus*, *B. psithyrus*, *B. pratorum*, and *B. jonellus*, the medium-tongued *B. hypnorum*, and the long-tongued *B. gerstaeckeri*. Models 2 and 3 both explained approximately 31% of total variance (97.5% CI = 30–32%).

@@

While bumble bees are, indeed, remarkably generalized in their foraging, they are highly discriminating generalists, dramatically over-representing certain floral taxa and morphotypes and ignoring others. In our study system, lip, flag, and head flowers dominated overall visitation probability when visitation was pooled across all bumble bee species, while bell, disc, and stalk-disc flowers were underrepresented. Patterns of pooled visitation are, of course, driven disproportionately by the behavior of the most abundant bumble bee species, which in our system were *B. pascuorum*, *B. pratorum*, *B. soroeensis*, *B. terrestris/lucorum*, and *B. wurflenii* (Sponsler et al., submitted). Visitation to lip and flag flowers was driven especially by the medium-tongued *B. pascuorum*, the most abundant bumble bee in our study system, and secondarily by the medium-tongued *B. wurflenii*. Visitation to head flowers was driven mainly by the short-tongued *B. soroeensis*.

Tongue length class was an important predictor of preference for some but not all floral morphotypes. For disc and stalk-disc flowers, visitation rates were uniformly low across all tongue length classes. The opposite was true of lip flowers, which were visited at high rates by all tongue length classes. Bell flowers and head flowers were preferred by short-tongued bees, consistent with the shallow corollas of these morphotypes. Conversely, funnel flowers, with their deep corollas were preferred by long-tongued bees, though this effect was driven disproportionately by *B. gerstaeckeri*, which specializes on the funnel flowers of the genus *Aconitum* [@Ponchau2006-af]. Flag flowers were visited at the highest rate by medium-tongued species, and especially by *B. pascuorum* and *B. mucidus*.

While the variation we observed in floral preference across tongue length classes largely comports with the established relationship between tongue length and corolla depth, variation in preference *within* tongue length classes echoes the findings of @Goulson2008-zy in demonstrating the insufficiency of simple morphological matching in explaining bumble bee foraging patterns. Perhaps the most striking divergence of floral preference within tongue length class was the divisive effect of bell flowers on short-tongued bumble bees. *B. pyrenaeus*, *B. soroeensis*, *B. pratorum*, and *B. jonellus* visited bell flowers at high rates, while *B. terrestris/lucorum*, *B. psithyrus*, and *B. lapidarius* visited them relatively rarely. A similar split among the short-tongued species occurred with respect to head flowers, which were visited at high rates by *B. soroeensis* and *B. psithyrus* but at relatively low rates by other short-tongued species. There are no obvious morphological distinctions within short-tongued bumble bees that would account for these divergent affinities, and ecological drivers such as differences in spatial or temporal activity patterns are more likely explanations [@Sowig1989-ld; @Schneller2014-ry]. Nectar robbing is another mechanism by which the relationship between tongue length and corolla depth might be relaxed, but its influence in our data is unclear. Of the bumble bees in our study system, *B. terrestris/lucorum* and especially *B. wurflenii* are known to engage frequently in nectar robbing [@Brian1954-cy]. In the case of *B. wurflenii*, however, its morphotype affinities did not differ markedly from those of other medium-tongued species, and it may be that the condition of medium-tongue length alone confers enough to foraging flexibility in our study system to dampen the divergent effects of nectar robbing. Nectar robbing might, however, account for the deviation of *B. terrestris/lucorum* from most other short-tongued species with respect to flag flowers. With the exception of *B. lapidarius*, all short-tongued species besides *B. terrestris/lucorum* eschewed flag flowers, while *B. terrestris/lucorum* visited them at rates comparable to medium- and long-tongued species. Indeed, B. terrestris/lucorum was, on the whole, an idiosyncratic forager, also demonstrating a relatively high affinity for the disc flowers that were largely ignored by other species. Together, these species-level idiosyncracies of preference within tongue length classes suggest a possible answer to the question of how bumble bees of similar tongue length can coexists, namely that factors other than tongue length are evidently at play in bumble bees' morphological affinities.

With respect to diet overlap among bumble bees, our finding that bumble bee diet ordination produces a primary axis corresponding to tongue length (and associated with affinity for legumes) and a secondary axis describing the dispersion of short-tongued species is consistent with studies in Poland [@Goulson2008-oe], France [@Durieux2000-uk], and England [@Goulson2005-nz] that analyzed bumble bee diet by similar methods, suggesting that this may be a general pattern among bumble bee communities, at least within temperate Europe. Perhaps a more significant finding, though, is that community-mean diet overlap at the level of floral morphotype was consistently nearly twice as high as diet overlap at the level of floral genus. From a numerical perspective, this result is unsurprising simply because each morphotype comprises many genera. From an ecological perspective, though, this pattern suggests another answer to the question of how bumble bee species of similar tongue length can coexists. If tongue length primarily constrains bumble bee foraging at the coarse level of floral morphotype, taxonomic diversity within morphotypes creates space for bumble bees with similar tongue lengths to avoid actual dietary overlap, thus mitigating interspecific competition. Importantly, this suggests that taxonomic diversity *within* floral morphotypes may as important as morphotype diversity *per se* and overall floral abundance in facilitating the coexistence of bumble bee species, an insight which may be relevant to applied bumble bee conservation.

With respect to elevation, community mean niche overlap was highest at the lowest sites and declined sharply over the transition from valley floor to mid-slope. If this pattern of niche overlap moves upslope in response to climate change, it might result in increased competition potential at higher elevations, which could threaten high-elevation bumble bee species that currently exist under conditions of relatively low niche overlap. In our system, though, it must be noted that the lowest elevation sites are also the ones most influences by human land use, and it is possible that the elevation pattern evident in our data is driven by factors other than elevational climate variation.

<!-- If not tongue length, what else could explain the variation in floral morphotype preference witin tongue length classes? -->

# CONCLUSION

The mystery of bumble bee species coexistence hinges on the relationship between species-level floral preferences, trait-matching, and the emergent phenomenon of community-level patterns of diet overlap. While tongue length is a relatively strong predictor of species-level floral preferences, it is at best a weak predictor of diet overlap. This is due in part to divergent preference of bumble bees with similar tongue lengths toward certain floral morphotypes. Perhaps more importantly, though, diet overlap at the level of floral genus tended to be about half as strong as diet overlap at the level of floral morphotype, indicating that bumble bees with similar morphological affinities, driven in part by tongue length compatibility, subdivide floral morphotypes at the genus- and species-level, reducing the potential for competition. This suggests that taxonomic floral diversity may be as important as overall floral abundance in enabling the coexistence of bumble bee species. Something about elevation...

@@

<!-- We analyze community mean niche overlap as an indication of diffuse competition [@Pianka1974-qk; MacArthur1972-qz] -->
